waRne - Putting cricket analytics in a spin

### 2020-10-02 CopyOfpaper_rob_working-.Rmd
cleaned up the text a bit, added the win prob for first inning tems

---
title: "waRne - Putting cricket analytics in a spin"
authors: Robert(1), James (2)
output: 
  pdf_document:
    toc: yes
  html_document:
    toc: true
    toc_float: true
    keep_md: true
bibliography: references.bib
---

```{r setup,message=FALSE, include=FALSE}
knitr::opts_chunk$set(message = TRUE, warning = TRUE)
library(tidyverse)
player_database <- 
  read_csv("data/new_player_database_crickjam (2).csv") %>% 
  dplyr::select(player_id, player_name, batting_style, bowling_style) %>% 
  dplyr::distinct() %>% 
  dplyr::filter(player_name != "Rohit Gurunath Sharma") %>% 
  bind_rows(read_csv("data/new_player_database_crickjam (2).csv") %>% 
              dplyr::filter(player_id == 34102) %>% 
              dplyr::select(player_id, player_name, batting_style, bowling_style) %>% 
              drop_na() %>% 
              distinct()) %>% 
  dplyr::filter(player_name !="Sultan Ahmed")

@@

cricjam_data <- 
  read_csv("data/combined_2020_09_11.csv") %>% 
  dplyr::filter(!grepl("No result", match_summary)) %>%
  mutate(bucketed_year = if_else(nchar(year)==4, year, as.numeric(str_sub(as.character(year), 1, 4)))) %>% 
  left_join(batting_vars, by = c("batsman_id" = "player_id")) %>% 
  left_join(bowler_vars, by = c("bowler_id" = "player_id")) %>% 
  mutate(batsman_name = str_remove(batsman_name, "\\(\\)"), 
         batsman_name = str_trim(batsman_name), 
         runs = extra_runs+earned_runs)
# 
# cricjam_data_partner <- 
#   cricjam_data %>% 
#   dplyr::select(match_id, inning, batsman_name, batsman_id, batting_order) %>% 
#   distinct() %>% 
#   arrange(batting_order) %>% View()

@@

The importance of reproducibility, and the related issue of open access to data, has received a lot of recent attention.\^[why this not work] Momentum on these issues is gathering in the sports analytics community. While cricket is the second largest commercial sport in the world, unlike other popular international sports, there has been no mechanism for the public to access comprehensive statistics on players and teams. Expert commentary currently relies heavily on data that isn't made readily accessible and this produces an unnecessary barrier for the development of an inclusive sports analytics community.

@@

By making data more accessible and more advanced metrics more accessible fans data journalism in sports has grown in recent years. For example in [@horowitz2017nflscrapr] has enabled EPA to enter popular discussion among fans.

@@

@@

Through being able to reproduce, fans are able to make accessible findings for others and importantly they are able to extend and grow concepts. Unfortuntely what we see through leading cricket analytics providers is a track record of confusing output for fans. This can lead to lower engagement and dismissial of cricket analytics.

For example in this [series of tweets](https://twitter.com/benjonescricket/status/1219898578719723520) we see the narrative being pushed that Steve Smith is a good player vs pace bowling unfortunetly just a few months prior the same company and journalist published an article which had Steve Smith doing much worse against pace (balls above 140km/h). Unlike a similar sport baseball, fans have no easily accisible way of seeing if Steve Smith vs pace is a strength as alluded to in the original tweet, or a weakness like the same persons published online article. In comparsion, fans are able to get a breakdown of Mike Trout vs fastballs from using [baseballr](https://billpetti.github.io/2018-02-19-build-statcast-database-rstats/) which provides access through statcast data from [baseballsavant](https://baseballsavant.mlb.com/).

Unfortunetly this is just one of many examples whereby a relatively simple statistic is provided by media and fans have no mechanism to fact check. Fact checking is an important avenue for fans to not only engage and understand statistics, but having this mechanism also stops analytics people/companies from putting out misleading conclusions and findings.

@@

@@

Fans and analysts of the game might want to make this decision based on a simple statistic, for example they might make it based on answer these questions instead.

-   Do teams that bat second have a higher winning percentage than those who bat first?

-   Is this consistent across leagues and levels of cricket?

@@

cricjam_data_big %>% 
	dplyr::filter(tournament_name %in% c("Twenty20 Big Bash", "Big Bash League"), 
								inning == 2) %>% 
	mutate(year = if_else(nchar(year)==4, year, as.numeric(str_sub(as.character(year), 1, 4)))) %>% 
	group_by(match_id) %>% 
	slice(1) %>% 
	dplyr::select(batting_team_result, year) %>% 
	mutate(binary_result = if_else(batting_team_result=="win", 1, 0)) %>% 
	group_by(batting_team_result, year) %>% 
	summarise(win_count = n()) %>% 
	ungroup() %>% 
	pivot_wider(names_from = batting_team_result, values_from = win_count) %>% 
	rename(superover = 'super-over') %>% 
	replace_na(list(superover=0)) %>% 
	mutate(total_games = lose+win+superover, 
				 win_percentage = win/total_games, 
				 year = as.factor(year)) %>% 
	ggplot(aes(x = year, y = win_percentage, group = 1)) +
	geom_line(color = 'red')
```

Instead of just looking at this graphically, fans of crickets and teachers of undergraduate statistics might use the dataset as a ''biased coin'' example. So instead of asking, given a proportion of heads over \$n\$ tosses instructors could ask is a team better off batting first or second.

@@

n<-900
p<-0.5
qbinom(c(0.025,0.975),size=n,prob=p)

@@

>>>>

---
title: "waRne - Putting cricket analytics in a spin"
authors: Robert(1), James (2)
output: 
  html_document:
    toc: true
    toc_float: true
    keep_md: true
  pdf_document:
    toc: yes
bibliography: references.bib
---

```{r setup,message=FALSE, include=FALSE}
knitr::opts_chunk$set(message = TRUE, warning = TRUE)
library(tidyverse)
player_database <- 
  read_csv("/Users/James/James/Cricket_Thesis/Data/new_player_database_crickjam.csv") %>% 
  dplyr::select(player_id, player_name, batting_style, bowling_style) %>% 
  dplyr::distinct() %>% 
  dplyr::filter(player_name != "Rohit Gurunath Sharma") %>% 
  bind_rows(read_csv("/Users/James/James/Cricket_Thesis/Data/new_player_database_crickjam.csv") %>% 
              dplyr::filter(player_id == 34102) %>% 
              dplyr::select(player_id, player_name, batting_style, bowling_style) %>% 
              drop_na() %>% 
              distinct()) %>% 
  dplyr::filter(player_name !="Sultan Ahmed")

@@

cricjam_data <- 
  read_csv("/Users/James/James/Reproducibility-in-Cricket/Data/combined_2020_09_11.csv") %>% 
  dplyr::filter(!grepl("No result", match_summary)) %>%
  mutate(bucketed_year = if_else(nchar(year)==4, year, as.numeric(str_sub(as.character(year), 1, 4)))) %>% 
  left_join(batting_vars, by = c("batsman_id" = "player_id")) %>% 
  left_join(bowler_vars, by = c("bowler_id" = "player_id")) %>% 
  mutate(batsman_name = str_remove(batsman_name, "\\(\\)"), 
         batsman_name = str_trim(batsman_name), 
         runs = extra_runs+earned_runs)
# 
# cricjam_data_partner <- 
#   cricjam_data %>% 
#   dplyr::select(match_id, inning, batsman_name, batsman_id, batting_order) %>% 
#   distinct() %>% 
#   arrange(batting_order) %>% View()

@@

The importance of reproducibility, and the related issue of open access to data, has received a lot of recent attention.\^[why this not work] Momentum on these issues is gathering in the sports analytics community. While cricket is the second largest commercial sport in the world, unlike other popular international sports, there has been no mechanism for the public to access comprehensive data on players and teams. Expert commentary currently relies heavily on data that isn't made readily accessible, and this produces an unnecessary barrier for the development of an inclusive sports analytics community.

@@

By making data more accessible and more advanced metrics more accessible to fans data journalism in sports has grown in recent years. For example in [@horowitz2017nflscrapr] has enabled EPA to enter popular discussion among fans.

@@

Unless otherwis specified, examples are done using data from 2010-2020 T20I, Big Bash, IPL, and Champions League cricket.

@@

Through being able to analyze the data ourselves, fans are able to make accessible findings for others and importantly they are able to extend and grow concepts. Unfortuntely what we see through leading cricket analytics providers is a track record of confusing output for fans. This can lead to lower engagement and dismissial of cricket analytics.

For example in this [series of tweets](https://twitter.com/benjonescricket/status/1219898578719723520) we see the narrative being pushed that Steve Smith is a good player vs pace bowling. Unfortunetly just a few months prior the same company and journalist published an article which had Steve Smith doing much worse against pace (balls above 140km/h). Unlike a similar sport baseball, fans have no easily accisible way of seeing if Steve Smith vs pace is a strength as alluded to in the original tweet, or a weakness like the same persons earlier article. In comparsion, fans are able to get a breakdown of Mike Trout vs fastballs from using [baseballr](https://billpetti.github.io/2018-02-19-build-statcast-database-rstats/) which provides access through statcast data from [baseballsavant](https://baseballsavant.mlb.com/).

Unfortunately this is just one of many examples whereby a relatively simple statistic is provided by media and fans have no mechanism to fact check. Fact checking is an important avenue for fans to not only engage and understand statistics, but having this mechanism also stops analytics people/companies from putting out misleading conclusions and findings.

@@

The data provided in the waRne package does not include the speed of the deliveries, but the type of ball bowled (and all available statistics and descriptions) can be taken from the personal player pages and joined into the main data set. A simple example would be to look at Steve Smiths performance against spin and pace bowling.

@@

Fans and analysts of the game might want to make this decision based on a simple statistic, for example they might make it based on answer a different question. Do teams that bat second have a higher winning percentage than those who bat first? Is this consistent across leagues and levels of cricket? We look at this graohically first

@@

cricjam_data_big %>% 
	dplyr::filter(tournament_name %in% c("Twenty20 Big Bash", "Big Bash League"), 
								inning == 2) %>% 
	mutate(year = if_else(nchar(year)==4, year, as.numeric(str_sub(as.character(year), 1, 4)))) %>% 
	group_by(match_id) %>% 
	slice(1) %>% 
	dplyr::select(batting_team_result, year, match_id) %>% 
	mutate(binary_result = if_else(batting_team_result=="win", 1, 0)) %>% 
	group_by(batting_team_result, year) %>% 
	summarise(win_count = n()) %>% 
	ungroup() %>% 
	pivot_wider(names_from = batting_team_result, values_from = win_count) %>% 
	rename(superover = 'super-over') %>% 
	replace_na(list(superover=0)) %>% 
	mutate(total_games = lose+win+superover, 
				 win_percentage = win/total_games, 
				 year = as.factor(year)) %>% 
	ggplot(aes(x = year, y = win_percentage, group = 1)) +
	geom_line(color = 'red') +
	theme_minimal()+
	labs(y = "Win Percentage", 
			 x = "Year")
```

Alternatively, fans of crickets and teachers of undergraduate statistics might use the dataset as a ''biased coin'' example. So instead of asking, given a proportion of heads over \$n\$ tosses instructors could ask is a team better off batting first or second. One thing that is often brought up my analysts and fans is: is there an advantage to batting second? We perform a simple hypothesis test on the proportion winning teams that bat first, and found that there was no significant evidence to support the notion that teams batting first win less than 50% of games.

@@

dat <- 
	cricjam_data_big %>% 
	dplyr::filter(tournament_name %in% c("Twenty20 Big Bash", "Big Bash League"), 
								inning == 2) %>% 
	mutate(year = if_else(nchar(year)==4, year, as.numeric(str_sub(as.character(year), 1, 4)))) %>% 
	group_by(match_id) %>% 
	slice(1) %>% 
	dplyr::select(batting_team_result, year, match_id) %>% 
	mutate(binary_result = if_else(batting_team_result=="win", 1, 0)) %>% 
	group_by(batting_team_result) %>% 
	summarise(win_count = n()) %>% 
	dplyr::filter(batting_team_result != 'super-over')

N = sum(dat$win_count)
wins = dat$win_count[2]

z = (wins/N-0.5)/sqrt((0.5^2)/N)

out <- pnorm(z)

@@

Probably drop this since the win prob is done now

### 2020-10-02 James_paper_copy.Rmd
cleaned up the text a bit, added the win prob for first inning tems

We recreated this example from the James-Stein Bookdown to show how cricket is similar to baseball in it's utility for teaching a wide range of statistical concepts. With a new open source data set available, the statistical community and fans have a rich new data set to work with for both teaching and as fans.

Take the James-Stein estimator as an example. We choose batsmen who have face d aminimum of 400 balls in the 2019 and 2019/2020 seasons, and using their final strike rate as the "true" strike rate, we estmate at various points in the season what their final strike rate will be. Since T20 cricket is played on the international stage, and at various domestic tournaments which players from around the world are allowed to play in, we will just choose the 10%, 25%, and 50% of their totals balls faced for the season as these estimation points.

@@

players <-
	cricjam_data_big %>% 
	group_by(batsman_id, batsman_name) %>% 
	# dplyr::filter(year %in% c("2019", "20192020")) %>%
	mutate(balls_faced = n()) %>% 
	dplyr::filter(balls_faced > 2500) %>% 
	distinct() %>% 
	# summarise(total = n())
	pull(batsman_id)

truth <- 
	cricjam_data_big %>% 
  filter(
  	# year %in% c("2019", "20192020"), 
  			 batsman_id %in% players) %>% 
  group_by(batsman_id) %>% 
  summarise(true_sr = 100*sum(extra_runs+earned_runs)/n(), 
              count_full = n()) %>% 
  select(batsman_id,true_sr, count_full)

set.seed(42)
obs <- 
	cricjam_data_big %>% 
  filter(
  	# year %in% c("2019", "20192020"), 
  			 batsman_id %in% players) %>% 
  group_by(batsman_id) %>%
  do(sample_frac(., 0.1))  %>%       
  summarise(mle_10 = 100*sum(extra_runs+earned_runs)/n(), 
  					count_10 = n()) %>% 
  select(batsman_id,mle_10, count_10) %>% 
  inner_join(truth,by="batsman_id") %>% 
	left_join(cricjam_data_big %>% 
  filter(
  	# year %in% c("2019", "20192020"), 
  			 batsman_id %in% players) %>% 
  group_by(batsman_id) %>%
  do(sample_frac(., 0.25))  %>%       
  summarise(mle_25 = 100*sum(extra_runs+earned_runs)/n(), 
  					count_25 = n()) %>% 
  select(batsman_id,mle_25, count_25), by = "batsman_id") %>% 
	left_join(cricjam_data_big %>% 
  filter(
  	# year %in% c("2019", "20192020"), 
  			 batsman_id %in% players) %>% 
  group_by(batsman_id) %>%
  do(sample_frac(., 0.50))  %>%       
  summarise(mle_50 = 100*sum(extra_runs+earned_runs)/n(), 
  					count_50 = n()) %>% 
  select(batsman_id,mle_50, count_50), by = "batsman_id")

@@

errors <- 
	obs %>% 
  mutate(sigma2 = sum((mle_10-p_10_)^2)/count_10,
         JS=p_10_+(1-((N-3)*sigma2/(sum((mle_10-p_10_)^2))))*(mle_10-p_10_)) %>% 
  mutate(mle_pred_error_i = (mle_10-true_sr)^2,
         js_pred_error_i = (JS-true_sr)^2) %>% 
  summarise(js_pred_error = sum(js_pred_error_i),
            mle_pred_error = sum(mle_pred_error_i))

@@

	errors_50 <- 
	obs %>% 
  mutate(sigma2 = sum((mle_50-p_50_)^2)/count_50,
         JS=p_50_+(1-((N-3)*sigma2/(sum((mle_50-p_50_)^2))))*(mle_50-p_50_)) %>% 
  mutate(mle_pred_error_i = (mle_50-true_sr)^2,
         js_pred_error_i = (JS-true_sr)^2) %>% 
  summarise(js_pred_error = sum(js_pred_error_i),
            mle_pred_error = sum(mle_pred_error_i))

>>>>

We recreated this example from the James-Stein Bookdown to show how cricket is similar to baseball in it's utility for teaching a wide range of statistical concepts. With a new open source data set available, the statistical community and fans have a rich new data set to work with for any number of personal or professional projects

We choose batsmen who have face a minimum of 400 balls in the 2019 and 2019/2020 seasons, and using their final strike rate as the "true" strike rate, we estmate at various points in the season what their final strike rate will be. Since T20 cricket is played on the international stage, and at various domestic tournaments which players from around the world are allowed to play in, we will just choose the 10%, 25%, and 50% of their totals balls faced for the season as these estimation points.

@@

players <-
	cricjam_data_big %>% 
	group_by(batsman_id, batsman_name) %>% 
	dplyr::filter(year %in% c("2019", "20192020")) %>%
	mutate(balls_faced = n()) %>% 
	dplyr::filter(balls_faced > 400) %>% 
	distinct(batsman_id) %>% 
	# summarise(total = n())
	pull(batsman_id)

truth <- 
	cricjam_data_big %>% 
  filter(
  	year %in% c("2019", "20192020"),
  			 batsman_id %in% players) %>% 
  group_by(batsman_id) %>% 
  summarise(true_sr = 100*sum(extra_runs+earned_runs)/n(), 
              count_full = n()) %>% 
  select(batsman_id,true_sr, count_full)

set.seed(42)
obs <- 
	cricjam_data_big %>% 
  filter(
  	year %in% c("2019", "20192020"),
  			 batsman_id %in% players) %>% 
  group_by(batsman_id) %>%
  do(sample_frac(., 0.1))  %>%       
  summarise(mle_10 = 100*sum(extra_runs+earned_runs)/n(), 
  					count_10 = n()) %>% 
  select(batsman_id,mle_10, count_10) %>% 
  inner_join(truth,by="batsman_id") %>% 
	left_join(cricjam_data_big %>% 
  filter(
  	year %in% c("2019", "20192020"),
  			 batsman_id %in% players) %>% 
  group_by(batsman_id) %>%
  do(sample_frac(., 0.25))  %>%       
  summarise(mle_25 = 100*sum(extra_runs+earned_runs)/n(), 
  					count_25 = n()) %>% 
  select(batsman_id,mle_25, count_25), by = "batsman_id") %>% 
	left_join(cricjam_data_big %>% 
  filter(
  	year %in% c("2019", "20192020"),
  			 batsman_id %in% players) %>% 
  group_by(batsman_id) %>%
  do(sample_frac(., 0.50))  %>%       
  summarise(mle_50 = 100*sum(extra_runs+earned_runs)/n(), 
  					count_50 = n()) %>% 
  select(batsman_id,mle_50, count_50), by = "batsman_id")

@@

errors_10 <- 
	obs %>% 
  mutate(sigma2 = sum((mle_10-p_10_)^2)/count_10,
         JS=p_10_+(1-((N-3)*sigma2/(sum((mle_10-p_10_)^2))))*(mle_10-p_10_)) %>% 
  mutate(mle_pred_error_i = (mle_10-true_sr)^2,
         js_pred_error_i = (JS-true_sr)^2) %>% 
  summarise(js_pred_error = sum(js_pred_error_i),
            mle_pred_error = sum(mle_pred_error_i))

@@

	errors_50 <- 
	obs %>% 
  mutate(sigma2 = sum((mle_50-p_50_)^2)/count_50,
  			 c = (1-((N-3)*sigma2/(sum((mle_50-p_50_)^2)))),
         JS=p_50_+c*(mle_50-p_50_)) %>% 
  mutate(mle_pred_error_i = (mle_50-true_sr)^2,
         js_pred_error_i = (JS-true_sr)^2) %>% 
  summarise(js_pred_error = sum(js_pred_error_i),
            mle_pred_error = sum(mle_pred_error_i))
